An application of Graph Theory to MDAO problem formulation

### 2012-10-01 problem_formulation.tex
DJP: added to the graph syntax section

\documentclass[]{aiaa-tc} % insert '[draft]' option to show overfull boxes

 \title{An application of Graph Theory to MDAO problem formulation}

\author{
  David Pate, %
     \thanks{Georgia Tech}
  Dr. Brian German 
     \thanks{Georgia Tech...}
  Justin Gray,%
     \thanks{Aerospace Engineer, MDAO Branch, Mail Stop 5-11, AIAA Member}   
 }

%\usepackage{setspace}
%\doublespace

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{caption} 
\usepackage{amsmath}
\usepackage{lscape}
\usepackage{hyperref}
\usepackage{appendix}
\usepackage[section]{placeins}

\captionsetup[figure]{margin=5pt,font=small,labelfont=bf,textfont=bf,justification=justified,}
%\captionsetup[wrapfigure]{margin=5pt,font=small,labelfont=bf,justification=justified,singlelinecheck=off}
\captionsetup[table]{margin=5pt,font=small,labelfont=bf,textfont=bf,justification=justified,position=top}

\bibliographystyle{aiaa}

\usepackage{lettrine}
\usepackage{verbatim}

%\usepackage{hyperref} %allows for the creation of actual text links
\begin{document}

\maketitle

\begin{abstract}
   blah blah blah
\end{abstract}

\section*{Nomenclature}

\begin{tabular}{l l} 
    AAO      & All-At- \\
    MDAO     & Multidisciplinary Design Analysis and Optimization \\
    FPF      & Fundamental Problem Formulation \\
\end{tabular}

\section{Introduction}

    As the size and complexity of engineering systems grows the time and expense for setting up 
    analysis models grows with them. Multidisciplinary Design Analysis and Optimization (MDAO)
    frameworks such as OpenMDAO\cite{Gray2012} and ModelCenter have enabled a new level of analysis tool integration 
    and paved the way for models of with more analysis tools and increasing numbers of multidisciplinary couplings. 
    Such complex models present a distinct challenge proper implementation of any given solution strategy . 
    In fact as the size of a problem grows very large even determining what the proper solution strategy can be a daunting 
    task.

    What is needed, in order to address the complexity problem, is a syntax for specifying an engineering problem 
    in a way that allows for both analysis of potential solution strategies and automated implementation of those strategies 
    inside an MDAO framework. In order to serve that purpose, the problem formulation needs to include the following information: 
    \begin{itemize}
       \item Discipline Analyses 
       \item Discipline State Variables and Residuals
       \item Local \& Global Design Variables
       \item Local \& Global Constraints
       \item Objective or Objectives
       \item Coupling Constraints
       \item Local and Global Parameters
    \end{itemize}
    Notably absent from the preceding list are any kind of solvers, optimizers, or other iterative solution finding tools. 
    At it most basic, a problem formulation includes only information about what is being sought after in a given problem, or what the
    goals of a given problem are. It need not contain any information about solution paths or strategies to reach those goals. 
    A general specification is one that states nothing specific about a problem solution path.

    We define a complete and general problem specification striped down to it's most essential parts to be the 
    Fundamental Problem Formulation (FPF). By definition, the FPF for any given problem will be constant regardless 
    of which MDAO framework, optimization architecture, optimizer, or solver is used to solve the problem.

    In this work we proposed a graph based syntax for the specification of problem formulation. This graph based syntax provides several key
    features that make it useful for working with large scale MDAO problems. It provides a rigid structure that can be easily manipulated 
    with a wide range of well establish graph-theory algorithms for the purposes of problem decomposition. The graph syntax also 
    provides a means for algorithmically testing a given problem formulation to check if it is the FPF, and if not, to reduce it to the FPF. 
    Lastly, a graph based specification for problem formulation lends itself well to interacting with MDAO frameworks which dramatically 
    increases it's utility for real world applications.

\section{Problem Formulation Syntax}
    The most common problem formulation syntax for an MDAO problem could be given as follows:

    \begin{align}
        min &\ \ f\left(X\right) \notag
        \\ w.r.t. &\ \  X \notag
        \\ s.t. &\ \ G\left(X\right) \leq 0
        \\      &\ \ C\left(X\right) = 0
    \end{align}

    Where $f(X)$ is the objective function, $X$ is the vector of design variables, $G(X)$ are the optimization constraints, 
    and $C(X)$ are the coupling constraints. This does represent a general problem formulation, since it says nothing about 
    how the problem should be solved. Tedford and Martins used this format to specify the FPF for a set of test problems and 
    also to describe specific formulations for solving them with a number of optimization architectures\cite{Tedford2009}. Their
    work demonstrates clearly how multiple specific problem formulations can all relate back to a common FPF.

    The challenge with using this traditional mathematical syntax is that it is not easily manipulated or analyzed by computer programs. 
    A number of matrix based methods have been used successfully to translate the mathematical syntax into a more useful computational form. 
    Steward's Design Structure Matrix (DSM) is a form of an adjacency matrix which captures the relationship between analysis tools where off 
    diagonal elements of the matrix indicate coupling\cite{Steward1981}. Figure \ref{fig:dsm_simple} shows a simple notional example of a DSM. 
    The DSM has been successfully applied in a number of ways to algorithmically modify problem formulations to 
    improve computational efficiency. Rogers et. al developed DeMAID to manipulate a
    DSM  minimizing the computational costs of solving highly coupled systems\cite{Rogers1996}. Lu and Martins demonstrated 
    the use of a weighted DSM to make partitioning and coordinating choices that dramatically reduced the 
    cost of solving a specific kind of optimization problem as the scale of the problem grew. 
    They applied weights to both the nodes (diagonal terms) and the edges (off diagonal terms) of the 
    DSM to indicated computational cost and coupling strength respectively.

    \begin{figure}[!hbp]
        \begin{center}
        \includegraphics[width=.75\textwidth]{images/dsm_simple}
        \caption{Graph and matrix based DSM specifications for a notional analysis \label{fig:dsm_simple}}
        \end{center}
    \end{figure}

@@

\section{Graph Based Problem Formulation Syntax}
    \subsection{Formulation Graph Syntax}
    Rather than start with an adjacency, we chose to work directly with a directed cyclic graph to develop a syntax for the FPF. 
    \begin{itemize}
        \item Specify analyses
        \item Connections between analyses 
            \begin{itemize}
                \item local variables
                \item global variables? Use "fake" node that broadcasts out to the rest of the graph? 
            \end{itemize}
        \item Cycles indicate coupling
        \item Cycles for design variables->objectives/constraints
        \item Objectives/constraints are just outputs? Special nodes? 
        \item Residuals are just outputs? Special nodes? 
        \item Parameters are just input nodes that are not design variables (use identifies these)
        \item FPF no solvers/optimizers anywhere in it
    \end{itemize}

>>>>

@@

\section{Graph Based Problem Formulation Syntax}
    \subsection{Formulation Graph Syntax}
    Rather than start with an adjacency, we chose to work directly with a directed cyclic graph to develop a syntax for the FPF. 
    The following information must be provided to start:
    \begin{itemize}
        \item Analysis blocks: an analysis block represents any calculation, and each comes with
            \begin{itemize}
                \item local inputs
                \item local outputs 
                \item execution properties: the properties associated with running the code, such as run time.
            \end{itemize}
        \item Global parameters: these may serve as fixed inputs to the local inputs of analysis blocks
        \item Global outputs: these may represent
            \begin{itemize}
                \item objectives
                \item constraints
                \item residuals: (not sure)
            \end{itemize}
    \end{itemize}

    The graph representation of a data flow is cast to utilize the extensive library of algorithms in graph theory to analyze a directed weighted graph. 
    Edge weights are used to represent the metrics associated with a data flow
    \begin{itemize}
        \item Run time: this metric is a property of an analysis block
        \item Fidelity: this metric is a property of an individual local output
        \item Expected Convergence
    \end{itemize}

    The key assumption is that identical variables are recognized as such. This serves as the basis for creating a data flow by connecting compatible input and output nodes with a directed edge. 
    In order to represent the fact that execution of an analysis code does not depend on the number of outputs being used, it is represented as depicted in the following figure (not made yet).

    This information immediately leads to the maximal connectivity graph, which is formed by placing a directed edge from each local output or global parameter to each matching local input or global output. 
    Whenever multiple edges are connected to a single input a conflict occurs because only one may be used. Resolving these conflicts is one key challenge in creating a data flow.
%    \begin{itemize}
%        \item Specify analyses
%        \item Connections between analyses 
%            \begin{itemize}
%                \item local variables
%                \item global variables? Use "fake" node that broadcasts out to the rest of the graph? 
%            \end{itemize}
%        \item Cycles indicate coupling
%        \item Cycles for design variables->objectives/constraints
%        \item Objectives/constraints are just outputs? Special nodes? 
%        \item Residuals are just outputs? Special nodes? 
%        \item Parameters are just input nodes that are not design variables (use identifies these)
%        \item FPF no solvers/optimizers anywhere in it
%    \end{itemize}
